---
title: "Assignment 5: Data Visualization"
author: Carina Huang
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(here)
library(cowplot)
getwd()

peterpaul1 <- read.csv(here("Data/Raw/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)

niwo <- read.csv(here("Data/Raw/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"),
         stringsAsFactors = TRUE)

#2 

peterpaul1$sampledate <- ymd(peterpaul1$sampledate)
niwo$collectDate <- ymd(niwo$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"),
        axis.title = element_text(family = 'Avenir'),
        plot.title = element_text(family = 'Avenir', size = 18),
        legend.position = "top",
        legend.title = element_text(family = 'Avenir')
        )

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
plot4 <- ggplot(peterpaul1, aes(x=tp_ug, y=po4, color = lakename))+
  geom_point()+
  geom_smooth(method = lm, se=FALSE)+
  ylim(0,60)+
  mytheme+
  labs(title = "Peter and Paul Lake phosphorus and phosphate level")+
  labs(x= "time")
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

plot_temp <- ggplot (peterpaul1, aes(x=month, y=temperature_C, color = lakename))+
  geom_boxplot()+
  aes(x=factor(
      month,
      levels = 1:12,
      labels = month.abb
  ))+
  labs(x= "Month")+
  theme(legend.position = "none")
plot_tp <- ggplot (peterpaul1, aes(x=month, y=tp_ug, color = lakename))+
  geom_boxplot()+
   aes(x=factor(
      month,
      levels = 1:12,
      labels = month.abb
  ))+
  labs(x= "Month")+
  theme(legend.position = "none")

plot_tn <- ggplot (peterpaul1, aes(x=month, y=tn_ug, color = lakename))+
  geom_boxplot()+
   aes(x=factor(
      month,
      levels = 1:12,
      labels = month.abb
  ))+
  labs(x= "Month")

legend <- get_legend(plot_tn)

plot_grid(plot_temp, plot_tp, plot_tn+theme(legend.position = "none"),legend,
          nrow = 4, align = 'h')


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: during warmer months (seasons), the lake temperature is higher, 
and the different nutrient levels are also higher compared to colder months.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
plot_needle <-filter(niwo, functionalGroup == "Needles")
drymass_plot <- ggplot(niwo, aes(x=factor(collectDate), 
                                 y=dryMass, color = nlcdClass))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle =45))
  
print(drymass_plot)
#7
drymass_plot_facet <- ggplot(niwo, aes(x=factor(collectDate), y=dryMass))+
  geom_boxplot()+
 facet_wrap(facets = vars(nlcdClass),
             nrow =3)+
  theme(axis.text.x = element_text(angle =45, vjust = 1))+
  labs(x=NULL)


print(drymass_plot_facet)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:I think plot 6 is more effective because it is easier to compare the 
dry mass of the different categories on the same plot collected on the same 
day. 